---
title: "Project_1_Group_1"
output: html_document
date: "2024-09-29"
names: "Jayden Benzan, Rebecca Beneroff, Kayla Lichtner"
---
Tutorial: How to determine which statistical tests to use for your analysis?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Tutorial introduction 


##Tutorial objectives
The purpose of this tutorial is to lead one through the process of deciding the suitable statistical analysis for their given dataset. Specifically, we will discuss when it's appropriate to use t-tests, anova testing, and linear modeling. There are three objectives for this tutorial:

1. Understand how to classify data (understanding different variable types and its effect on how to use them in analysis).
2. Review loading data, organizing data, and preparing data for statistical analysis.
3. Explain the conditions required for using linear modeling, t.tests, or ANOVA tests to analyze data. 


### Instructions

## Loading the necessary packages 

```{r}
if (!require("UsingR")) install.packages("UsingR"); library(UsingR) # For the simple.eda function
if (!require("cowplot")) install.packages("cowplot"); library(cowplot) # For graphing aesthetics
if (!require("conflicted")) install.packages("conflicted"); library(conflicted) # For conflicts
if (!require("readxl")) install.packages("readxl"); library(readxl) #To read in data
if (!require("gplots")) install.packages("gplots"); library(gplots) #For graphs/plots
if (!require("UsingR")) install.packages("UsingR"); library(UsingR) #Generally good to have!
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse) # Always download this!
conflict_prefer_all("dplyr", quiet = TRUE)
```

## Importing the dataset
Go to our Git repository called Project_1_Group_1_2024, and download the "iris.extended.csv" data file from the "Dataset options" folder. We are going to make a final folder with the finished tutorial and the data csv file, but for now, please do this to import the data.
```{r}
iris.raw.data <- read.csv("iris_extended.csv")
```

Citation for the dataset:

After loading in a dataset, you always want to make sure that everything loaded in properly. To do this, check the top and bottom of your data, and the number of observations you have imported. The data on GitHub reports 1200 lines of data, in our environment the data should read 1200 observations as there is one line for column headings.

```{r}
head(iris.raw.data)
tail(iris.raw.data)
nrow(iris.raw.data)
```
This shows us that there are 1200 rows of data which aligns with the expectation of 1200 lines of data according to Github.

It is important to understand the structure of the raw data as it will impact how you prepare for analysis. To do so:
```{r}
glimpse(iris.raw.data)
```

## Looking into different variables- Jayden 


## Exploratory Data Analysis
It is important to run an initial exploratory data analysis to understand the data distributions as the distribution of data directly influences the type of statistical analysis to perform. 

To explore the distributions of some of the variables we are interested in:

```{r}

```








## Acknowledgements

